Delivering Reliable Software in the European Scientific Arena: Evolution of Methodologies and New Trend Analysis

### 2018-02-06 paper-ieee-trans.tex
CD description improved

In this paper, we introduce a set of EC-funded projects
that go from 2001 up to now, in which the authors of this summary were involved.
For each project, we describe the challenges arised, the methodologies used and
the achievements towards software reliability. The reminder of this paper is
organized as follows. Section \ref{sec:ev} details the evolution of research on
software reliability in EC-projects. Section \ref{sec:ntsr} provides
information about new research trends in software reliability based upon the 
experiences from the recent INDIGO-DataCloud project. Finally, Section \ref{sec:con} concludes.

@@

\begin{table}[!h]
%% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{List of EC-projects}
\label{tab:eup}
\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
\begin{tabular}{p{1.6cm}p{1.5cm}p{3cm}l}
\hline
\hline
\\
Logo & Short Name & Long Name & Duration\\
\hline
\hline
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/datagrid}
\end{minipage}
    & DataGrid &
Research and Technological Development for an International Data Grid & 2001--2003\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egee}
\end{minipage}
     & EGEE I, II, III &
Enabling Grids for E-sciencE & 2004--2010\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/etics}
\end{minipage}
     & ETICS 1, 2 &
E--Infrastructure for Testing, Integration and Configuration of Software & 2006-2010\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/emi}
\end{minipage}
     & EMI &
Europeean Middleware Initiative & 2010--2013\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egi-inspire}
\end{minipage}
     & EGI-Inspire &
Integrated Sustainable Pan-European Infrastructure for Researchers in Europe
 & 2010--2014\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egi_engage}
\end{minipage}
     & EGI Engage &
Engaging the EGI Community towards an Open Science Commons
 & 2015--2017\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/indigo}
\end{minipage}
     & INDIGO-DataCloud &
INtegrating Distributed data Infrastructures for Global ExplOitation
 & 2015--2017\\
\hline
\hline
\end{tabular}
\end{table}

@@

\begin{table}[!h]
%% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{Features on software reliability in EC-projects}
\label{tab:feat}
\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
\begin{adjustbox}{max width=0.5\textwidth}
\begin{tabular}{llllllll}
\hline
\hline
Feature & DataGrid & EGEEs & ETICSs & EMI & EGIs & INDIGO-DC\\
\hline
\hline
Architecture Task Force&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Communication Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Requirements Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Agile Methodologies&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Source code inspection&&&&$\surd$&&$\surd$\\
Build/Testing Management Procedure&$\surd$&$\surd$&$\surd$&$\surd$&&$\surd$\\
Software Product Metrics&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Quality Criteria Definition&&&$\surd$&$\surd$&$\surd$&$\surd$\\
%Manual Certification Procedure&$\surd$&$\surd$&&$\surd$&&\\
Automatic Certification&&&$\surd$&&$\surd$&$\surd$\\
Auto-generated Documentation&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
DevOps practices adoption (CI, CD)&&&&&&$\surd$\\
\hline
\hline
\end{tabular}
\end{adjustbox}
\end{table}

@@

The INDIGO-DataCloud (INtegrating Distributed data Infrastructures for Global
explOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the last
project of software development considered in this review. At the time of writing
this paper, the project is succeeding in providing solutions to address the existing gaps in
Cloud Platform--as--a--Service (PaaS) and Software--as--a--Service (SaaS) levels,
helping developers, e--Infrastructures and scientific communities to exploit
Cloud computing benefits.

@@

\item \textbf{Unit and Functional testing}.
Changes involving the addition of new functionalities are requested to be tested. \textit{Regression
testing} in this context is done enforcing the successful execution of the previous
functionality tests already present in the code, so to uncover any malfunction introduced by
the changes. On the other hand, \textit{unit testing} completes the source code testing evaluation by
focusing on the code's internal design. Project's SQA criteria sets a recommended threshold of
70\% coverage for unit testing. In Figure \ref{fig:fig_unittest}, unit testing coverage is compared
for each software component over the two major INDIGO-DataCloud releases, showing an increment in the 
averaged value of {\sl INDIGO-2} release. As it can be seen, a very small set of products with low coverage
values highly impacted the overall result, although a significant part of these products are upstream
contributions or products with codebases not fully under the control of the project. Nevertheless,
60\% of the products were over this threshold while 80\% of them had values above
the 50\% coverage.

@@

\item \textbf{Automated Deployment}.
In order to lower the barriers of software adoption, the SQA criteria requires the automated 
deployment of the products delivered as part of the catalogue. Automation in this context is 
tackled using a configuration management tool. These tools allow to have an 
\textit{infrastructure as code} (IaC), managing the component's deployment through declarative
definitions. The definition files appear as additional source of documentation, 
self-documenting code, as they sequentially guide the component's deployment process in multiple
platforms. In this regard, INDIGO--DataCloud project contributed to open--source IaC tools such 
as Ansible \cite{indigo-ansible} and Puppet \cite{indigo-puppet}. A representative example of 
such contribution are the \textit{50 roles} being developed from scratch, currently hosted in
the Ansible Galaxy portal \footnote{https://galaxy.ansible.com/indigo-dc/}. Figure 
\ref{fig:fig_confman} shows the number of products that offer an automated means for deployment.
It shows an increase after the {\sl INDIGO-1} release especially in the previous weeks of each
of the two major releases.
\end{itemize}

@@

The evaluation of the software quality is performed by measuring the values of
the metrics and Key Performance Indicators (KPIs) defined based upon the
ISO/IEC 9126 standard. These metrics cover the development, release and
maintenance phases of the software lifecycle. Development and release metrics
are obtained automatically from several sources, such as GitHub and Jenkins CI
service, and graphically displayed as GitHub pages using GrimoireLab framework
\cite{grimoirelab}. Maintenance and user support metrics are collected from the
different sources of data such as GGUS \cite{ggus} and Github Issues.

@@

Figure \ref{fig:fig_jenkins_CI_builds} shows the evolution, throughout the 
INDIGO-DataCloud's first and second releases, of the required test types builds since the
implementation of the CI infrastructure. Automated functional testing coverage were not
available for all the software stack, thus the associated number of builds are fewer. The
executions are not precisely aligned as checks can be manually triggered, through Jenkins
interface, useful for instance whenever a failed check needs to be rebuilt.

\subsubsection{Continuous Delivery}
As DevOps suggests, frequent releases positively affect the reliability of the
software. The software updates of INDIGO--DataCloud products taking place since
the second major release are passing through a Continuous Delivery (CD)
pipeline that adds the packaging of the software right after the successful execution of
the quality checks (as part of the CI). 
The steps defined within the CD pipeline differ whether the software is to be distributed via Docker
images (see Figure \ref{fig:fig_CD}) or via operating system’s packages, rpms or debs. In the latter case an extra
validation step is added, consisting in the product’s deployment using a
Configuration Management (CM) solution. The installation uses the packages
built in the previous step and uploaded to a testing
repository. In the case of Docker images, the CM is used to build the image
itself, installing and configuring the product before uploading it to the
DockerHub repository \cite{indigo-dockerhub}.

@@

\begin{figure*}	
	\centering
	\begin{subfigure}
		\centering
\includegraphics[width=0.85\textwidth]{images/disvis-flow.png}
\caption{DevOps pipeline to distribute Docker images for Disvis application.}
\label{fig:fig_disvis}		
	\end{subfigure}
	\quad
	\begin{subfigure}
		\centering
\includegraphics[width=\textwidth]{images/pilotpreview.png}
\caption{Resource centers supporting the Pilot Preview testbed and corresponding
set of deployed INDIGO-DataCloud components or services.}
\label{fig:fig_pilotpreview}
	\end{subfigure}
\end{figure*}

@@

The experience gathered throughout the project with regard to the adoption of
different DevOps practices is not only useful and suitable for the software related
to the core services in the INDIGO-DataCloud solution, but also applicable to the
development and distribution of the applications coming from the user communities.

@@

User application developers were provided with both a means to validate the
source code before merging and the creation of a new versioned Docker image,
automatically available in the INDIGO-DataCloud’s application repository.

@@

\bibitem{cordis:indigo}
\emph{INtegrating Distributed data Infrastructures for Global ExplOitation
(INDIGO-DataCloud)} project, European Community Research and Development
Information Service (CORDIS),
http://cordis.europa.eu/project/rcn/194882\_en.html

\bibitem{indigo-dockerhub}
\emph{INDIGO-DataCloud DockerHub repository},
https://hub.docker.com/u/indigodatacloud

\bibitem{indigo-d31}
\emph{Initial Plan for WP3} INDIGO-DataCloud Deliverable 3.1,
https://www.indigo-datacloud.eu/documents/initial-plan-wp3-d31

\bibitem{indigo-jenkins}
\emph{INDIGO-DataCloud Jenkins CI service},
https://jenkins.indigo-datacloud.eu:8080/

\bibitem{indigo-github}
\emph{INDIGO-DataCloud GitHub Source Code repository},
https://www.github.com/indigo-dc

\bibitem{indigo-gitbook}
\emph{INDIGO-DataCloud GitBook Documentation repository},
https://www.gitbook.com/@indigo-dc

\bibitem{indigo-puppet}
\emph{INDIGO-DataCloud PuppetForge repository},
https://forge.puppet.com/indigodc

\bibitem{indigo-ansible}
\emph{INDIGO-DataCloud Ansible Galaxy repository},
https://galaxy.ansible.com/indigo-dc/

@@

@@

% if you will not have a photo at all:
\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/pablo-orviz}}]{Pablo Orviz Fernandez}
is a computing researcher at CSIC (Consejo Superior de Investigaciones
Cientificas), holding an MSc in Scientific Computing from the University of
Cantabria. During the past years he has been devoted to software quality
assurance tasks within different European advanced computing research projects
such as INDIGO-DataCloud and EGI-Engage. Currently involved in EOSC-hub project
and participating in the software quality work packages of DEEP-HybridDataCloud
and eXtreme-DataCloud.
\end{IEEEbiography}

@@

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/DavideSalomoni}}]{Davide Salomoni}
is Director of Technology at the Italian National Institute for Nuclear Physics (INFN). He has 27 years of international experience in private and public environments related to distributed computing and communication technologies. He currently leads the Software Development and Distributed Systems department at CNAF, the INFN National Center dedicated to research and development on IT technologies, located in Bologna, Italy. He was Project Coordinator of INDIGO-DataCloud, a project funded by the EC Horizon2020, that developed innovative open source computing and storage solutions.
\end{IEEEbiography}

>>>>

In this paper, we introduce a set of EC-funded projects
that go from 2001 up to now, in which the authors of this summary were involved.
For each project, we describe the challenges arised, the methodologies used and
the achievements towards software reliability. The reminder of this paper is
organized as follows. Section \ref{sec:ev} details the evolution of research on
software reliability in EC-projects. Section \ref{sec:ntsr} provides
information about new research trends in software reliability based upon the 
experiences from the recent INDIGO--DataCloud project. Finally, Section \ref{sec:con} concludes.

@@

\begin{table}[!h]
%% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{List of EC-projects}
\label{tab:eup}
\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
\begin{tabular}{p{1.6cm}p{1.5cm}p{3cm}l}
\hline
\hline
\\
Logo & Short Name & Long Name & Duration\\
\hline
\hline
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/datagrid}
\end{minipage}
    & DataGrid &
Research and Technological Development for an International Data Grid & 2001--2003\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egee}
\end{minipage}
     & EGEE I, II, III &
Enabling Grids for E-sciencE & 2004--2010\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/etics}
\end{minipage}
     & ETICS 1, 2 &
E--Infrastructure for Testing, Integration and Configuration of Software & 2006-2010\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/emi}
\end{minipage}
     & EMI &
Europeean Middleware Initiative & 2010--2013\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egi-inspire}
\end{minipage}
     & EGI-Inspire &
Integrated Sustainable Pan-European Infrastructure for Researchers in Europe
 & 2010--2014\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egi_engage}
\end{minipage}
     & EGI Engage &
Engaging the EGI Community towards an Open Science Commons
 & 2015--2017\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/indigo}
\end{minipage}
     & INDIGO--DataCloud &
INtegrating Distributed data Infrastructures for Global ExplOitation
 & 2015--2017\\
\hline
\hline
\end{tabular}
\end{table}

@@

\begin{table}[!h]
%% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{Features on software reliability in EC-projects}
\label{tab:feat}
\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
\begin{adjustbox}{max width=0.5\textwidth}
\begin{tabular}{llllllll}
\hline
\hline
Feature & DataGrid & EGEEs & ETICSs & EMI & EGIs & INDIGO--DC\\
\hline
\hline
Architecture Task Force&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Communication Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Requirements Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Agile Methodologies&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Source code inspection&&&&$\surd$&&$\surd$\\
Build/Testing Management Procedure&$\surd$&$\surd$&$\surd$&$\surd$&&$\surd$\\
Software Product Metrics&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Quality Criteria Definition&&&$\surd$&$\surd$&$\surd$&$\surd$\\
%Manual Certification Procedure&$\surd$&$\surd$&&$\surd$&&\\
Automatic Certification&&&$\surd$&&$\surd$&$\surd$\\
Auto-generated Documentation&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
DevOps practices adoption (CI, CD)&&&&&&$\surd$\\
\hline
\hline
\end{tabular}
\end{adjustbox}
\end{table}

@@

The INDIGO--DataCloud (INtegrating Distributed data Infrastructures for Global
explOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the last
project of software development considered in this review. At the time of writing
this paper, the project is succeeding in providing solutions to address the existing gaps in
Cloud Platform--as--a--Service (PaaS) and Software--as--a--Service (SaaS) levels,
helping developers, e--Infrastructures and scientific communities to exploit
Cloud computing benefits.

@@

\item \textbf{Unit and Functional testing}.
Changes involving the addition of new functionalities are requested to be tested. \textit{Regression
testing} in this context is done enforcing the successful execution of the previous
functionality tests already present in the code, so to uncover any malfunction introduced by
the changes. On the other hand, \textit{unit testing} completes the source code testing evaluation by
focusing on the code's internal design. Project's SQA criteria sets a recommended threshold of
70\% coverage for unit testing. In Figure \ref{fig:fig_unittest}, unit testing coverage is compared
for each software component over the two major INDIGO--DataCloud releases, showing an increment in the 
averaged value of {\sl INDIGO--2} release. As it can be seen, a very small set of products with low coverage
values highly impacted the overall result, although a significant part of these products are upstream
contributions or products with codebases not fully under the control of the project. Nevertheless,
60\% of the products were over this threshold while 80\% of them had values above
the 50\% coverage.

@@

\item \textbf{Automated Deployment}.
In order to lower the barriers of software adoption, the SQA criteria requires the automated 
deployment of the products delivered as part of the catalogue. Automation in this context is 
tackled using a configuration management tool. These tools allow to have an 
\textit{infrastructure as code} (IaC), managing the component's deployment through declarative
definitions. The definition files appear as additional source of documentation, 
self-documenting code, as they sequentially guide the component's deployment process in multiple
platforms. In this regard, INDIGO--DataCloud project contributed to open--source IaC tools such 
as Ansible \cite{indigo-ansible} and Puppet \cite{indigo-puppet}. A representative example of 
such contribution are the \textit{50 roles} being developed from scratch, currently hosted in
the Ansible Galaxy portal \footnote{https://galaxy.ansible.com/indigo-dc/}. Figure 
\ref{fig:fig_confman} shows the number of products that offer an automated means for deployment.
It shows an increase after the {\sl INDIGO--1} release especially in the previous weeks of each
of the two major releases.
\end{itemize}

@@

The evaluation of the software quality is performed by measuring the values of
the metrics and Key Performance Indicators (KPIs) defined based upon the
ISO/IEC 9126 standard. These metrics cover the development, release and
maintenance phases of the software lifecycle.

\textit{Development metrics} are obtained programatically from several sources, namely GitHub
API and Jenkins service, and graphically displayed as GitHub pages using GrimoireLab
framework \cite{grimoirelab}. Per-component weekly reports, including the SQA requirement
completeness, are issued and individually discussed through the different communication channels
with the development teams. Issue tracking metrics and KPIs are an essential piece of information
of both feature addition and defect solving, useful to detect and fix misbehaviours while in the
development phase.

\textit{Release metric} sources are mainly the online repository servers, namely the 
Linux package \cite{indigo-pkg-repo} and DockerHub \cite{indigo-dockerhub} repositories. Jenkins 
server contains as well valuable release data as it is the service where the packages
and containers are being built before being uploaded to the online repositories. Release
KPIs primarly focus on the frequency and effiency -- mainly rollbacks -- statistics.

\textit{Maintenance and user support metrics} are key for continuously improving the response
to issues reported by external users. Feeback is collected from outside helpdesks, such as 
EGI's GGUS \cite{ggus}, and the INDIGO--DataCloud's GitHub Issues tracker.

@@

Figure \ref{fig:fig_jenkins_CI_builds} shows the evolution, throughout the 
INDIGO--DataCloud's first and second releases, of the required test types builds since the
implementation of the CI infrastructure. Automated functional testing coverage were not
available for all the software stack, thus the associated number of builds are fewer. The
executions are not precisely aligned as checks can be manually triggered, through Jenkins
interface, useful for instance whenever a failed check needs to be rebuilt.

\subsubsection{Continuous Delivery}
As DevOps suggests, frequent releases positively affect the reliability of the software as
they involve fewer changes and usually more correlated. The software updates of 
INDIGO--DataCloud products, taking place since the second major release, are passing through
a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
successful execution of the CI pipeline previously described. Consequently, CI-validated 
software components are automatically delivered in online repositories without the need of 
any human interaction but ultimately requiring expert supervision to validate the results.

INDIGO--DataCloud software stack is delivered in form of Linux software packages (rpms, debs) 
and Docker containers. The CD approach is different for each type of software packaging. \textit{Linux
package pipeline} intentionally delivers the packages produced in a pre-production -- preview --
branch. Before that, once the software packages have been created, they are uploaded to a 
testing branch. The available automated deployment solution -- see \cite{subsec:sqa} -- checks
the installation and configuration of the component relying on the testing repository. On 
successful completion, the packages are then automatically signed and subsequently moved to 
the preview branch, from where they finally be made available in the production branch after
the release manager supervises the process.

The \textit{Docker container pipeline} uses instead the automated deployment solution to 
actually install and configure the software component in the container image. Once this is 
done the recently created image is uploaded to the production DockerHub repository, tagged
with the label corresponding to the current INDIGO-DataCloud release. Figure 
\ref{fig:fig_CD} shows the complete workflow of the Docker container CD implementation, 
automatically triggered by a change in the source code.

@@

\begin{figure*}	
	\centering
	\begin{subfigure}
		\centering
\includegraphics[width=0.85\textwidth]{images/disvis-flow.png}
\caption{DevOps pipeline to distribute Docker images for Disvis application.}
\label{fig:fig_disvis}		
	\end{subfigure}
	\quad
	\begin{subfigure}
		\centering
\includegraphics[width=\textwidth]{images/pilotpreview.png}
\caption{Resource centers supporting the Pilot Preview testbed and corresponding
set of deployed INDIGO--DataCloud components or services.}
\label{fig:fig_pilotpreview}
	\end{subfigure}
\end{figure*}

@@

The experience gathered throughout the project with regard to the adoption of
different DevOps practices is not only useful and suitable for the software related
to the core services in the INDIGO--DataCloud solution, but also applicable to the
development and distribution of the applications coming from the user communities.

@@

User application developers were provided with both a means to validate the
source code before merging and the creation of a new versioned Docker image,
automatically available in the INDIGO--DataCloud’s application repository.

@@

\bibitem{cordis:indigo}
\emph{INtegrating Distributed data Infrastructures for Global ExplOitation
(INDIGO--DataCloud)} project, European Community Research and Development
Information Service (CORDIS),
http://cordis.europa.eu/project/rcn/194882\_en.html

\bibitem{indigo-dockerhub}
\emph{INDIGO--DataCloud DockerHub repository},
https://hub.docker.com/u/indigodatacloud

\bibitem{indigo-d31}
\emph{Initial Plan for WP3} INDIGO--DataCloud Deliverable 3.1,
https://www.indigo-datacloud.eu/documents/initial-plan-wp3-d31

\bibitem{indigo-jenkins}
\emph{INDIGO--DataCloud Jenkins CI service},
https://jenkins.indigo-datacloud.eu:8080/

\bibitem{indigo-github}
\emph{INDIGO--DataCloud GitHub Source Code repository},
https://www.github.com/indigo-dc

\bibitem{indigo-gitbook}
\emph{INDIGO--DataCloud GitBook Documentation repository},
https://www.gitbook.com/@indigo-dc

\bibitem{indigo-puppet}
\emph{INDIGO--DataCloud PuppetForge repository},
https://forge.puppet.com/indigodc

\bibitem{indigo-ansible}
\emph{INDIGO--DataCloud Ansible Galaxy repository},
https://galaxy.ansible.com/indigo-dc/

@@

\bibitem{indigo-pkg-repo}
\emph{INDIGO--DataCloud Linux Package repository}, https://jenkins.indigo-datacloud.eu/

@@

% if you will not have a photo at all:
\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/pablo-orviz}}]{Pablo Orviz Fernandez}
is a computing researcher at CSIC (Consejo Superior de Investigaciones
Cientificas), holding an MSc in Scientific Computing from the University of
Cantabria. During the past years he has been devoted to software quality
assurance tasks within different European advanced computing research projects
such as INDIGO--DataCloud and EGI-Engage. Currently involved in EOSC-hub project
and participating in the software quality work packages of DEEP-HybridDataCloud
and eXtreme-DataCloud.
\end{IEEEbiography}

@@

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/DavideSalomoni}}]{Davide Salomoni}
is Director of Technology at the Italian National Institute for Nuclear Physics (INFN). He has 27 years of international experience in private and public environments related to distributed computing and communication technologies. He currently leads the Software Development and Distributed Systems department at CNAF, the INFN National Center dedicated to research and development on IT technologies, located in Bologna, Italy. He was Project Coordinator of INDIGO--DataCloud, a project funded by the EC Horizon2020, that developed innovative open source computing and storage solutions.
\end{IEEEbiography}
